---
title: "An Introduction to Functional Analysis"
author: "Alex Sanchez"
date: "April 1st 2021"
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, cache=FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE, results ='markup')
options(warn=-1)
```

```{r installation}
installifnot <- function (packageName){
 if (!(require(packageName, character.only=TRUE))) {
    install.packages(packageName)
  }else{
    detach(paste ("package", packageName, sep=":"), character.only=TRUE)
  } 
}
bioCifnot <- function (packageName){
 if (!(require(packageName, character.only=TRUE))) {
    BiocManager::install(packageName)
 }else{
  detach(paste ("package", packageName, sep=":"), character.only=TRUE)
  }  
}
installifnot("knitr")
installifnot("XML") # May yield problems if some libraries (xml2-config) not available in linux
bioCifnot ("org.Hs.eg.db")
bioCifnot ("hgu133a.db")
bioCifnot ("GO.db")
bioCifnot ("GOstats")
bioCifnot ("topGO")
bioCifnot ("GSA")
bioCifnot ("annotate")
bioCifnot ("Rgraphviz")
```


# Introduction

This document provides *some examples* on the analyses that can be perfomed on one or more gene lists to help gain biological insight on the results of a differential expression analysis. Overall these analyses are known as _Functional Analysis_.

Functional analysis can be performed in many different ways that lead to similar (or not-so-similar) results.  Because there is not a universal acceptance of what is a *complete, well done functional analysis* some different approaches will be shown.

## Input Data for Functional Analysis

Functional analysis can be made, on a first approach on

- A list of genes selected by being differentially expressed in a given experimental setting.
- The whole list of genes -or even the whole expression matrix- that has been used in the analysis.

Most tools require that gene list consist of gene identifiers in some standard notation such as `Entrez`, `ENSEMBL` or other related to these.

These gene lists can be usually extracted from output tables provided by microarrays or RNA-seq data analysis tools.

The analysis below is applied on a set of three gene lists obtained from a cancer study, but it can be easily extended to more lists or other studies.

## Read data

We start by reading two files that contain the expression values (`expres_AvsB.csv2`) and the results (`Top_AvsB.csv2`) of a differential expression analysis erformed using microarrays.

The code and text for the analysis that, using these data, generated these tables, can be found at: [https://github.com/alexsanchezpla/scripts/tree/master/Exemple_Analisis_BioC](https://github.com/alexsanchezpla/scripts/tree/master/Exemple_Analisis_BioC)

The code below assumes the files have been stored in a subdirectory of the current folder named `datasets`.

```{r readData1}
inputDir="datasets"
topTabAvsB <- read.table (file.path(inputDir, "Top_AvsB.csv2"), head=T, sep=";", dec=",", row.names=1)
expresAvsB <- read.table (file.path(inputDir, "expres_AvsB.csv2"), head=T, sep=";", dec=",", row.names=1)

dim(topTabAvsB); head(topTabAvsB)
dim(expresAvsB); head(expresAvsB)
```

# Exploring gene lists


A given gene list contains useful information that can be extracted by querying databases.

Let us see how we can obtain information fom the _genes_ probes in table (comparison) `AvsB`. 

```{r probes}
myProbes <- rownames(expresAvsB)
head(myProbes)
```


We need to load the library ("package") that contains specific annotations for the microarray model that was used in this study.
 It has to be noticed also that each row does not represent a gene, but a probe, a sequence that has been designed to detect if a gene is expressed. microarrays contain multiple probes for many genes and this is something that ha to be dealt with.
 
## ID conversion

In order to do most analyses it is convenient to use a universally accepted identifier such as Entrez or ENSEMBL.
For instance Bioconductor organism annotation packages rely on Entrez identifiers as main key for most mappings.

It is possible to find out easily which mappings are available for each ID

```{r mappings}
require(hgu133a.db)
keytypes(hgu133a.db)
```

Annotation packages make it possible to annotate genes and in a similar manner other omics features. For example we can obtain gene symbol, entrez ID and gene name with a single SQL instruction.
```{r}
geneAnots <- select(hgu133a.db, myProbes, c("SYMBOL", "ENTREZID", "GENENAME"))
head(geneAnots)
```

Now we can provide a more informative list of differentially expressed genes in topTable

```{r}
selected<- topTabAvsB[,"adj.P.Val"]<0.05 & topTabAvsB[,"logFC"] > 1
sum(selected)
selectedTopTab <- topTabAvsB[selected,]
selectedProbes <- rownames(selectedTopTab)
selectedAnots <-  select(hgu133a.db, selectedProbes, c("SYMBOL", "ENTREZID", "GENENAME"))
selectedInfo <- cbind(selectedAnots, selectedTopTab)
write.csv2(selectedInfo, file="selectedTopTab_AvsB.csv2")
```


<!--PER AFEGIR:  
- Una forma senzilla de treure una taula amb els IDs dels gens hiperenllaçats 
-->

## From gene lists to pathway analysis

See [http://ithake.eu/2012/09/on-the-exploration-of-a-list-of-genes/](This post) for a simple, external, motivation on why we may want to move from gene-by-gene analysis to gene-set analyses.


# Basic GO Annotation

<!-- ALTRES OPCIONS
Es poden trobar informacions de com consultar la GO a altres llocs:
Per exemple
- El meu vell document OntologyAnalysis.Rnw
- Al workflow d'anotació de Bioconductor: http://bioconductor.org/help/workflows/annotation/annotation/#OrgDb
- https://www.biostars.org/p/81174/ i els enllaços derivats

-->

Bioconductor libraries allow for both: 
- Exploration of functional information on genes based on the Gene Ontology
- Different types of Gene Enrichment and Pathway Analysis based on the GO or other pathway databases such as the KEGG, Reactome etc.

The code below shows some easy ways to retrieve GO information associated with genes

Start by loading the appropriate packages

```{r}
require(GOstats)
require(GO.db)
require(hgu133a.db); 
require(annotate) # Loads the required libraries.
```

Select the "top 5" genes from the list

```{r top25}
probes <- rownames(expresAvsB)[1:5]
```
For simplicity retrieve its Gene Symbol and Entrez identifiers using predefined functions ("wrappers") available in the `annotate` package.

```{r}
require(annotate)
entrezs <- unlist(mget (probes, hgu133aENTREZID))
simbols <- unlist(mget (probes, hgu133aSYMBOL))
```

Now proceed to retrieve GO information on these probes:
```{r}
GOAcc<-mget(probes,env=hgu133aGO)
GOAcc[[1]][1:5]
```

You might
This is a rather complicated list, indeed it is a list of lists of lists, that is

1. One list for each gene/ID containing
2. One list for each GO annotation for this gene containing
3. One list with the GOID, The Ontology and the Evidence Code

Indeed the third-level list can also be considered as an instance of a class defined in the `annotate` package: the class `Goterm`. Type `? GOTerm`
to know more about this class.

The following code will perform a wise unlist of the `GOAcc` list

<!-- ```{r} -->
<!-- x<-0 -->
<!-- for (i in 1:length(GOAcc)){x<-x+length(GOAcc[[i]])} -->
<!-- numGOterms <- x -->
<!-- res<- matrix(data = NA, nrow = numGOterms, ncol = 6) -->
<!-- actualrow<-1 -->
<!-- for (i in 1:length(GOAcc)){ -->
<!--     if (GOAcc[i]=="NA"){ -->
<!--       reg<-c(names(GOAcc[i]), entrezs[i], simbols[i], rep("NA",3)) -->
<!--     res[actualrow,]<-unlist(reg) -->
<!--     actualrow<-actualrow+1} -->
<!--     else{ -->
<!--           for (j in 1:length(GOAcc[[i]])){ -->
<!--             reg<-c(names(GOAcc[i]), entrezs[i], simbols[i],  -->
<!--                 GOAcc[[i]][[j]]$GOID, -->
<!--                 GOAcc[[i]][[j]]$Ontology, -->
<!--                 GOAcc[[i]][[j]]$Evidence) # or Evidence if preferred -->
<!--             res[actualrow,]<-unlist(reg) -->
<!--             actualrow<-actualrow+1}}} -->
<!-- rownames(res)=rep(NULL,dim(res)[1]) -->
<!-- res<-as.data.frame(res) -->
<!-- names(res)<-c("Affy-ID","Entrez", "Gene Symbol", "GOID","Ontology","Evidence") -->
<!-- print(head(res)) -->
<!-- ``` -->

```{r GOtable}
# % WANING
#  The previous chunk can be substituted by the followink code chunk, shorter and more efficient.
library(hgu133a.db)
keytypes(hgu133a.db)
res <- select(hgu133a.db, keys=probes, keytype = "PROBEID", columns = c("ENTREZID", "SYMBOL","ONTOLOGY"))
res1 <- select(hgu133a.db, keys=probes, keytype = "PROBEID",  columns = c("ENTREZID", "SYMBOL","GO"))
```

The resulting tables can be printed or saved.
```{r echo=TRUE}
print(head(res, n=10))
print(head(res1, n=10))
```


## Navigating the GO Graph

There are a series of classes to navigate up or down in the GO hierarchy
Take the top GO Term of the previous analysis: 
```{r}
oneTerm <- "GO:0006338"
oneParent<-get(oneTerm, GOBPPARENTS)   # the vector of its parent terms in the BP ontology.
oneParent
oneChildren<-get(oneTerm, GOBPCHILDREN) # the vector of its children terms in the BP ontolog
oneChildren
oneOffspring<-get(oneTerm, GOBPOFFSPRING) # the vector of its offspring terms in the BP ontology.
oneOffspring
oneChildren %in% oneOffspring
```

Similar analyses can be done on gene lists

```{r}
require(org.Hs.eg.db) # loads the library
myEIDs3 <-entrezs[1:3] # Create vecotor of input Entrez IDs
myGO <- unlist(org.Hs.egGO[[as.character(myEIDs3[1])]])
myGO_All <- mget(myEIDs3, org.Hs.egGO)
GOgenes <- org.Hs.egGO2ALLEGS[[myGO[1]]]
GOgenes_All <- mget(myGO[1], org.Hs.egGO2ALLEGS)
```


# Gene Enrichment Analysis
  
A gene enrichment analysis is applied on a "truncated" list of genes that one considers to be differentially expressed.

These are checked for enrichment versus a "Universe" gene list, usually, all the genes that have entered in the analysis

```{r}
require(hgu133a.db)
topTab <- topTabAvsB 
entrezUniverse = unlist(mget(rownames(topTab), hgu133aENTREZID, ifnotfound=NA)) 
whichGenes<- topTab["adj.P.Val"]<0.05 & topTab["logFC"] > 1
sum(whichGenes)
topGenes <-   entrezUniverse[whichGenes]
allMyGenes <-topTab$adj.P.Val
names(allMyGenes)<-rownames(topTab)
```

The analysis is performed by creating hyperparameters that are used to invoke the analysis functions
```{r}
require(topGO)
data(geneList) # adds function "topDiffGenes"
myGOData <- new("topGOdata", ontology="BP", 
                allGenes=allMyGenes,
                geneSel=topDiffGenes, nodeSize=10,  
                annot= annFUN.db, affyLib="hgu133a.db")

Myenrichment_Fisher <- runTest(myGOData, algorithm= "classic", statistic="fisher")
Myenrichment_Fisher

head(score(Myenrichment_Fisher), 25) # Displays p values for every GO term
geneData(Myenrichment_Fisher) # A table showing Medata data for enrichment
```

The topGO package implements different algorithms for gene set analysis.

```{r}
Myenrichment_KS <- runTest(myGOData, algorithm= "classic", statistic="ks")
```

Results can be shown in tabular format
```{r}
enrich_table <-GenTable(myGOData, classicFisher=Myenrichment_Fisher,topNodes = 20)
adjustedEnrichP <- cbind(enrich_table, adjP=p.adjust(enrich_table$classicFisher, method = "BH"))
head(adjustedEnrichP, n=25) # get the enrichment results as table
```

Different visualizations are also possible

```{r}
showSigOfNodes(myGOData, score(Myenrichment_Fisher), firstSigNodes=5, useInfo="all") # Plot the enrichment GO graph
gostat <- termStat(myGOData, names(score(Myenrichment_Fisher)))
plot(score(Myenrichment_Fisher), score(Myenrichment_KS)[names(score(Myenrichment_Fisher))], xlab="P values Fisher test", ylab="P values KS test", cex=(gostat$Annotated/max(gostat$Annotated))*4, col=heat.colors(gostat$Significant))
print(showGroupDensity(myGOData, enrich_table[1, "GO.ID"], ranks=TRUE))
```

# Gene Set Enrichment Analysis

If, instead of relying on the gene lists we decided to use all the genes on the array and confront them to _selected sets of genes_ we may use the *Gene Set Enrichment Analysis* approach.

There are many GSA implementations. We present here an applicarion of Efron's GSA method (http://statweb.stanford.edu/~tibs/GSA/).

```{r, eval=FALSE}
require(GSA)
xAB <- expresAvsB
yAB <- substr(colnames(expresAvsB),1,1)
yABnum <- ifelse(yAB=="A",1,2)
allProbes <- rownames(expresAvsB)
allEntrezs <- as.integer(select(hgu133a.db, allProbes, "ENTREZID")[,2])
allSymbols <- select(hgu133a.db, allProbes, "SYMBOL")
geneset.obj <-GSA.read.gmt("h.all.v5.2.entrez.gmt")
GSA.obj<-GSA(xAB, yABnum, genenames=allEntrezs, genesets=geneset.obj,  
             resp.type="Two class unpaired", nperms=200)
```
For simplicity retrieve its Gene Symbol and Entrez identifiers using predefined functions ("wrappers") available in the `annotate` package.

```{r, eval=FALSE}
require(annotate)
entrezs <- unlist(mget (probes, hgu133aENTREZID))
simbols <- unlist(mget (probes, hgu133aSYMBOL))
 Provar amb 1000?
  sink()
  cat("\nConjuntos de genes que aparecen en la comparacion: ",matNames[i],"\n")
  cat("=====================================================================","\n")  
  show(GSA.listsets(GSA.obj, geneset.names=genesetsNames, FDRcut=.5))
}
```


