---
title: "An Introduction to Functional Analysis"
author: "Alex Sanchez"
date: "April 4th 2021"
output:
  html_document:
    theme: united
    toc: yes
  pdf_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, cache=FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE, results ='markup')
options(warn=-1)
```

```{r installPackages, eval=FALSE}
installifnot <- function (packageName){
 if (!(require(packageName, character.only=TRUE))) {
    install.packages(packageName)
  }else{
    detach(paste ("package", packageName, sep=":"), character.only=TRUE)
  } 
}
bioCifnot <- function (packageName){
 if (!(require(packageName, character.only=TRUE))) {
    BiocManager::install(packageName)
 }else{
  detach(paste ("package", packageName, sep=":"), character.only=TRUE)
}  
}
installifnot("knitr")
installifnot("XML") # May yield problems if some libraries (xml2-config) not available in linux
bioCifnot ("org.Hs.eg.db")
bioCifnot ("hgu133a.db")
bioCifnot ("GO.db")
bioCifnot ("GOstats")
bioCifnot ("topGO")
bioCifnot ("annotate")
bioCifnot ("Rgraphviz")
bioCifnot ("clusterProfiler")
```


# Introduction

This document provides *some examples* on the analyses that can be perfomed on one or more gene lists to help gain biological insight on the results of a differential expression analysis. Overall these analyses are known as _Pathway  Analysis_ or, also, _Functional Analysis_. 

Functional analysis can be performed in many different ways that lead to similar (or not-so-similar) results.  Because there is not a universal acceptance of what is a *complete, well done functional analysis* some different approaches will be shown.

## Input Data for Functional Analysis

Functional analysis can be made, on a first approach on:

- One or more lists of genes __selected__ for being differentially expressed in a given experimental setting. In this case we usually work with gene identifiers.
- One or more list of values measuring the difference between groups (i.e. Fold Changes, p-values, or t-statistics) for all genes being  compared.

Most tools require that gene list consist of gene identifiers in some standard notation such as `Entrez`, `ENSEMB` or other related to these.

These gene lists can be usually extracted from output tables provided by microarrays or RNA-seq data analysis tools.

The analysis below is applied on a set gene lists obtained from a cancer study, but it can be easily extended to more lists or other studies.

## Read data

We start by reading two files that contain the expression values (`expres_AvsB.csv2`) and the results (`Top_AvsB.csv2`) of a differential expression analysis performed using microarrays.

The code and text for the analysis that, using these data, generated these tables, can be found at: [https://github.com/ASPteaching/Ejemplo_de_Analisis_de_Microarrays_con_Bioconductor](https://github.com/ASPteaching/Ejemplo_de_Analisis_de_Microarrays_con_Bioconductor)

The code below assumes the files have been stored in a subdirectory of the current folder named `datasets`.

```{r readData1}
inputDir="datasets"
topTabAvsB <- read.table (file.path(inputDir, "Top_AvsB.csv2"), head=T, sep=";", dec=",", row.names=1)
expresAvsB <- read.table (file.path(inputDir, "expres_AvsB.csv2"), head=T, sep=";", dec=",", row.names=1)
comparisonName <- "AvsB"
dim(topTabAvsB); head(topTabAvsB)
dim(expresAvsB); head(expresAvsB)
```

# Exploring gene lists


A given gene list contains useful information that can be extracted by querying databases.

Let us see how we can obtain information fom the _probesets_ in table (comparison) `AvsB`. 

```{r probes}
myProbes <- rownames(expresAvsB)
head(myProbes)
```

We need to load the library ("package") that contains specific annotations for the microarray type that was used in this study.
 It has to be noticed also that each row _does not represent a gene, but a probeset_, a sequence that has been designed to detect if a given gene is expressed. Microarrays contain multiple probesets for many genes and this is something that ha to be dealt with.
 
## ID conversion

In order to do many analyses it is convenient to use a universally accepted identifier such as `Entrez` or `ENSEMBL`.
For instance Bioconductor organism annotation packages rely on `Entrez` identifiers as main key for most mappings.

It is possible to easily find out which mappings are available for each ID.

```{r mappings0}
library(hgu133a.db)
keytypes(hgu133a.db)
```

Annotation packages make it possible to annotate genes and in a similar manner other omics features. For example, we can obtain gene symbol, entrez ID and gene name with a single SQL instruction.
```{r}
geneAnots <- select(hgu133a.db, myProbes, c("SYMBOL", "ENTREZID", "GENENAME"))
head(geneAnots)
```

Now we can provide a more informative list of differentially expressed genes in topTable

```{r}
selected<- topTabAvsB[,"adj.P.Val"]<0.05 & topTabAvsB[,"logFC"] > 1
sum(selected)
selectedTopTab <- topTabAvsB[selected,]
selectedProbes <- rownames(selectedTopTab)
selectedAnots <-  select(hgu133a.db, selectedProbes, c("SYMBOL", "ENTREZID", "GENENAME"))
selectedInfo <- cbind(selectedAnots, selectedTopTab)
write.csv2(selectedInfo, file="selectedTopTab_AvsB.csv2")
```


<!--PER AFEGIR:  
- Una forma senzilla de treure una taula amb els IDs dels gens hiperenllaçats 
-->

## From gene lists to pathway analysis

Pathway Analysis is an extensive field of research and application and the aim of this document is not to summarize it but simply to illustrate some applications with R.

See (https://github.com/ASPteaching/An-Introduction-to-Pathway-Analysis-with-R-and-Bioconductor/raw/main/Slides/Intro2PathwayEnrichmentAnalysis-SHORT.pdf) for an introduction to the topic.

The most used database in Pathway Analysis is the Gene Ontology, which, as suggested by its name, is not a dabase but an ontology. The `GO.db` can be accessed using the same syntaxis
 as with gene identifiers.

# Basic GO Annotation

<!-- ALTRES OPCIONS
Es poden trobar informacions de com consultar la GO a altres llocs:
Per exemple
- El meu vell document OntologyAnalysis.Rnw
- Al workflow d'anotació de Bioconductor: http://bioconductor.org/help/workflows/annotation/annotation/#OrgDb
- https://www.biostars.org/p/81174/ i els enllaços derivats

-->

Bioconductor libraries allow for both: 
- Exploration of functional information on genes based on the Gene Ontology
- Different types of Gene Enrichment and Pathway Analysis based on the GO or other pathway databases such as the KEGG, Reactome etc.

The code below shows some easy ways to retrieve GO information associated with genes

Start by loading the appropriate packages

```{r}
require(GOstats)
require(GO.db)
require(hgu133a.db); 
require(annotate) # Loads the required libraries.
```

Select the "top 5" genes from the list

```{r top25}
probes <- rownames(expresAvsB)[1:5]
```

In the first versions of Bioconductor, identifiers were managed using environments and functions such as `get` or `mget`. While it still works, nowadays it has been substituted by the use of the `select` function which allows for a muach easier retrieving of identifiers.

`Select` returns a data.frame. If we need a character vector we can obtain it using an ad-hoc function sucha as the simple`charFromSelect`below.

```{r}
charFromSelect <- function(df,posNames=1, posValues=2){
  res <- df[,posValues]
  names(res) <- df[,posNames]
  return(res)
}
  
require(annotate)
geneIDs <-  select(hgu133a.db, probes, c("ENTREZID", "SYMBOL"))
entrezs <-  charFromSelect(geneIDs, 1, 2)
simbols <-  charFromSelect(geneIDs, 1, 3) 
```




```{r GOtable}
# % WANING
#  The previous chunk can be substituted by the followink code chunk, shorter and more efficient.
library(hgu133a.db)
keytypes(hgu133a.db)
res <- select(hgu133a.db, keys=probes, keytype = "PROBEID", columns = c("ENTREZID", "SYMBOL","ONTOLOGY"))
res1 <- select(hgu133a.db, keys=probes, keytype = "PROBEID",  columns = c("ENTREZID", "SYMBOL","GO"))
```

The resulting tables can be printed or saved.
```{r echo=TRUE}
print(head(res, n=10))
print(head(res1, n=10))
```

### What do these GO:XXXXXXX mean?

We can query the GO.db package


```{r mappings}
library(GO.db)
keytypes(GO.db)
```

```{r}
queryTerms <- res1$GO[1:3]
select(GO.db, queryTerms, c("GOID", "ONTOLOGY","TERM" ))
```


## Navigating the GO Graph

There are a series of classes to navigate up or down in the GO hierarchy
Take the top GO Term of the previous analysis: 
```{r}
oneTerm <- "GO:0006338"
oneParent<-get(oneTerm, GOBPPARENTS)   # the vector of its parent terms in the BP ontology.
oneParent
oneChildren<-get(oneTerm, GOBPCHILDREN) # the vector of its children terms in the BP ontolog
oneChildren
oneOffspring<-get(oneTerm, GOBPOFFSPRING) # the vector of its offspring terms in the BP ontology.
oneOffspring
oneChildren %in% oneOffspring
```

Similar analyses can be done on gene lists

```{r}
require(org.Hs.eg.db) # loads the library
myEIDs3 <-entrezs[1:3] # Create vecotor of input Entrez IDs
myGO <- unlist(org.Hs.egGO[[as.character(myEIDs3[1])]])
myGO_All <- mget(myEIDs3, org.Hs.egGO)
GOgenes <- org.Hs.egGO2ALLEGS[[myGO[1]]]
GOgenes_All <- mget(myGO[1], org.Hs.egGO2ALLEGS)
```


# Gene Enrichment Analysis

There are two main types of enrichment analysis:

- _Over-Representation Analysis_ takes a list of differentially expressed genes and it searches for biological categories in which a number of genes appear with "unusually" high frequencies. That is it looks for genes appearing more often than they would be expected by chance in any category.

- _Gene Set Expression Analyses_ works with __all__ genes and looks for differentially expressed gene sets (categories). That is it searches for categories that, without containing an unusually high number of differentially expressed genes, are associated with genes that are in the upper or lower parts of the list of genes ordered by some measure of intensity of difference, such as the "log-Fold Change".

## Over-Representation Analysis 

Over-Representation Analysis is applied on a "truncated" list of genes that one considers to be differentially expressed.

These are checked for enrichment versus a "Universe" gene list, usually, all the genes that have entered in the analysis

```{r}
require(hgu133a.db)
topTab <- topTabAvsB 
probesUniverse <- rownames(topTab)
# entrezUniverse = unlist(mget(rownames(topTab), hgu133aENTREZID, ifnotfound=NA)) 
entrezUniverse<- select(hgu133a.db, probesUniverse, "ENTREZID")
entrezUniverse <- entrezUniverse$ENTREZID
whichGenes<- topTab["adj.P.Val"]<0.05 & topTab["logFC"] > 1
sum(whichGenes)
topGenes <-   entrezUniverse[whichGenes]

# Remove possible duplicates

topGenes <- topGenes[!duplicated(topGenes)]
entrezUniverse <- entrezUniverse[!duplicated(entrezUniverse)]
```

Many packages can be used to do a Gene Enrichment Analysis. Each of them perfoms slightly different analysis but the underlying ideas are the same. Some of these packages are:

- `GOstats`
- `topGO`
- `gprofiler`
- `clusterProfiler`

In this example we use `GOstats` which was one of the first packages available in Bioconduictor to do Enrichment Analysis.

The idea for using it is pretty simple: We need to create special type of object called "Hyperparameter" which may be of class either:

- `GOHyperGParams` if we want to do the enrichment analysis using the Gene Ontology as reference databases or
- `KEGGHyperGParams` if we wish to do the analysis using the KEGG database, or
- `PFAMHyperGParams` if we wish to do the analysis using the PFAM database.

A good strategy is to use all three.

First, we create the hyper-parameters.

```{r}
library(GOstats)

# This parameter has an "ontology" argument. It may be "BP", "MF" or "CC"
# Other arguments are taken by default. Check the help for more information.

GOparams = new("GOHyperGParams",
    geneIds=topGenes, universeGeneIds=entrezUniverse,
    annotation="hgu133a.db", ontology="BP",
    pvalueCutoff=0.001)

# These hyperparameters do not have an "ontology" argument.

KEGGparams = new("KEGGHyperGParams",
    geneIds=topGenes, universeGeneIds=entrezUniverse,
    annotation="hgu133a.db",
    pvalueCutoff=0.001)

PFAMparams = new("PFAMHyperGParams",
    geneIds=topGenes, universeGeneIds=entrezUniverse,
    annotation="hgu133a.db", 
    pvalueCutoff=0.001)
```

Next, we run the analyses:

```{r}
GOhyper = hyperGTest(GOparams)
KEGGhyper = hyperGTest(KEGGparams)
PFAMhyper = hyperGTest(PFAMparams)
```

We can extract information directly from the analysis results using `summary`:

```{r}
head(summary(PFAMhyper))
dim(summary(KEGGhyper))
```

Apart of this we can create an automated report per each result, or, with a litle more effort we can combine all outputs into a single one.


```{r}
# Creamos un informe html con los resultados
comparison = "AvsB"
GOfilename =file.path(paste("GOResults.",comparison,".html", sep=""))
htmlReport(GOhyper, file = GOfilename, summary.args=list("htmlLinks"=TRUE))
```

The resulting file has the contents provided by the call to `summary`but with additional links.


## Gene Set Enrichment Analysis

If, instead of relying on the gene lists we decided to use all the genes on the array and confront them to _selected sets of genes_ we may use the *Gene Set Enrichment Analysis* approach.

### Classical GSEA

The `clusterProfiler` package implements the classical GSEA method as introduced by Subramanian et alt (2005).


```{r}
entrezIDs <- AnnotationDbi::select(hgu133a.db, rownames(topTabAvsB), c("ENTREZID"))
entrezIDs <- charFromSelect(entrezIDs)
geneList <- cbind(topTabAvsB, ENTREZ =entrezIDs)

#ordenem per logFC absolut per eliminar els duplicats amb menor logFC absolut
geneList <- geneList[order(abs(geneList$logFC), decreasing=T),]
geneList <- geneList[ !duplicated(geneList$ENTREZ), ]  ### Keep highest
    #tornem a ordenar per logFC per fer el GSEA
geneList <- geneList[order(geneList$logFC, decreasing=T),]
genesVector <- geneList$logFC
names(genesVector) <- geneList$ENTREZ


#fixem seed per reproduibilitat dels resultats
#  set.seed(123)
library(clusterProfiler)
gseResulti <- gseKEGG(geneList = genesVector,
                      organism = "hsa",
                      keyType = "kegg",
                      exponent = 1,
                      minGSSize = 10,maxGSSize = 500,
                      pvalueCutoff = 0.05,pAdjustMethod = "BH",
                      # nPerm = 10000, #augmentem permutacions a 10000
                      verbose = TRUE,
                      use_internal_data = FALSE,
                      seed = TRUE,
                      eps=0,
                      by = "fgsea"
                )

# keggResultsList[[i]] <- gseResulti
```


```{r results='asis'}
library(kableExtra)
gsea.result <- setReadable(gseResulti, OrgDb = org.Hs.eg.db, keyType ="ENTREZID" )

gsea.result.df <- as.data.frame(gsea.result)
print(kable(gsea.result.df[,c("Description","setSize","NES","p.adjust")])%>% scroll_box(height = "500px"))
  
```


```{r eval=TRUE}
library(ggplot2)
# for (i in 1:length(files)){
#   cat("\nComparison: ", namesC[i],"\n")
   cat("DOTPLOT\n")
#   if(nrow(keggResultsList[[i]]) > 0){
 if(nrow(gseResulti) > 0){
   p<- dotplot(gseResulti, showCategory = 20, font.size = 15,
            title =paste("Enriched Pathways\n", comparisonName ,
            split=".sign") + facet_grid(.~.sign))
   plot(p)
   cat("\nENRICHMENT MAP\n")
   em<- emapplot(gseResulti)
   plot(em)
   #guardem en pdf
   pdf(file = paste0("KEGGplots.",comparisonName,".pdf"), 
                        width = 14, height = 14)
   print(p)
   print(em)
   dev.off()
   }else{
      cat("\nNo enriched terms found\n")
 }

```

